Typed Interactions for NLU in Opal

### 2017-12-13 paper.tex
Changes to formalism section
- Added nicer text for function definitions
- Revisions of wording, etc

@@

Finally, writing specific entity declarations rather than just \ts{type}, makes
it clear that there is more going on than a direct translation into typescript.
In fact, the DSL parser does lot of validation to guarantee that only reasonable
types can be written down. For example, a \ts{keywords} declaration is required
to be a union, and that union must contain only literals. Even the \ts{alias}
keyword has some validation built in. Some Typescript types would be problematic
if they were allowed---we enforce those conditions in the DSL.

@@

We elected to use a typescript library called \emph{runtypes} to serialize the
AST directly into typescript; this gives us a data structure to guide response
parsing, and also provides us a way to dynamically type-check the results. We
discuss the response parsing algorithm in more detail in section
\ref{implementation}.

@@

\section{Formalism} \label{formalism}
In order to reason about the correctness of our translations, we have formalized
the results from section \ref{description}. We begin by formalizing the type DSL
as a pseuo-BNF grammar. This is shown in figure \ref{fig:grammar_a}.
\begin{figure}
  \centering
  \begin{subfigure}{1\linewidth}
    \begin{align*}
      \fcy{D} ::=&\ \text{FreeText}(\etag{t}) \\
      |&\ \text{Keywords}(\etag{t},\ [\ell]) \\
      |&\ \text{Trait}(\etag{t},\ \{s: \fcy{T}\}) \\
      \\
      \fcy{T} ::=&\ \fcy{D} \mid \ell \mid \{s: \fcy{T}\}
    \end{align*}
    $$ \etag{t} \in \textsf{Tag} \qquad \ell, s \in \textbf{string} $$
    \caption{Type DSL}
    \label{fig:grammar_a}
  \end{subfigure}\vspace{1cm}
  \begin{subfigure}{1\linewidth}
    $$ \tau ::=\ \textbf{string} \mid \ell \mid \{s: \tau\} \mid \langle \ell
    \rangle $$
    \caption{Relevant Opal Types}
    \label{fig:grammar_b}
  \end{subfigure}
  \caption{Formal grammars.}
\end{figure}
There are similarities between these definitions and the haskell ones in figure
\ref{fig:ast}, but it has some slightly different goals. First, we remove
aliases and require that all type definitions are written ``in-line''. Next,
rather than split the definition by allowing tags within $\fcy{T}$, we also
require that entitiy definitions are inlined. This means that the data a user
expects from the NLU backend can be represented by a single tree of type
$\fcy{D}$. Importantly, none of these simplifications are fundamental---they
simply remove some aspects of the system that make it easier to use in practice.

@@

As mentioned above, user's view of their system can be formalized as a top-level
declaration generated by $\fcy{D}$. Since this declaration encompases both type
and configuration information, we would like a way to compute just the
\emph{type} of data that the user expects from the NLU engine. We define the
following mutually recursive functions, which traverse a declaration and
construct a single type in $\tau$.
\begin{align*}
  \text{interp}\ &:\ \fcy{D} \to \tau \\
  \text{interpTy}\ &:\ \fcy{T} \to \tau
\end{align*}

@@

\subsection{Configuration Generation}
The first step that our system takes is configuration. In practice, this is a
two-part process: the user needs to extract the appropriate Opal code as well as
the configuration for the NLU backend. The first step is covered by our
definition of interp. Since we are taking $\tau$ as an analog for relevant Opal
types, the type generation process is exactly the execution of interp on the
top-level entity declaration.

As for the second step, we would like the NLU backend to be set up such that its
responses are \emph{well-formed} with respect to the configuration. Intuitively,
any time we look up some entity tag in the response, we would like to get data
that is compatible with that entity's declaration. We define a type,
\textsf{Resp} of responses, and require that a lookup function exists, with the
following type:
$$ \text{lookup}\ :\ \textsf{Tag} \to \textsf{Resp} \rightharpoonup
\textbf{string} $$
Note that the function is partial; there is no guarantee that the response will
contain the desired data. In the sequel, we will often assume that lookup is
defined on tags we care about, but that assumption will be stated explicitly
rather than assumed by default.

The lookup function always returns a string, but the string might mean any
number of things. The data might be a particular literal, in the case of a
keywords or trait entity, or an entire phrase in the case of a free-text entity.
We define configurations as functions that map a subset of entity tags to the
type of data expected for those entities.
$$ \textsf{Cfg} \equiv \textsf{Tag} \rightharpoonup \tau $$
Now, we call a response well-formed with respect to $\text{cfg} : \textsf{Cfg}$
if and only if it returns appropriate data on lookup. More formally, for some $r
: \textsf{Resp}$,
\begin{align*}
  &r\ \text{well-formed}_{\text{cfg}} \iff \\
  &\forall (\etag{t} : \textsf{Tag}).\ \text{cfg}(\etag{t})\ \text{defined} \Rightarrow \text{lookup}(\etag{t}, r) : \text{cfg(\etag{t})}
\end{align*}

@@

Our goal will be to write a parsing function which takes a declaration, $d$, and
a response, and produces a value of type interp($d$). We will again write two
mutually recursive functions, parse and parseTy, with types:
\begin{align*}
\text{parse}\ &:\ \Pi_{(d : \fcy{D})}. \textsf{Resp} \rightharpoonup \text{interp}(d) \\
\text{parseTy}\ &:\ \Pi_{(t : \fcy{T})}. \textsf{Resp} \rightharpoonup \text{interp}(t)
\end{align*}
Since we need the type of the result to depend on the declaration, these are
dependent functions. The functions are also partial because lookups might fail.

We will define parse by case analysis on the declaration.
\begin{align*}
  \text{parse}(\text{FreeText}(\etag{t}), r) = \text{lookup}(\etag{t}, r)
\end{align*}
This case is fairly straightforward. Since the interpretation of a free-text
entity is just a string, we can simply look up the value of the appropriate
entity from the response.
\begin{align*}
  \text{parse}(\text{Keywords}(\etag{t}, [\ell_1, \dots, \ell_n]), r) =\\
  \text{lookup}(\etag{t}, r)
\end{align*}
Perhaps surprisingly, we can use the exact same definition in the Keywords case
that we did in the FreeText case. We are required to return a value of type
$\langle {\ell_1 \mid \dots \mid \ell_n} \rangle$; as long as $r$ is
well-formed, it will have been configured to have the same list of literals. We
are guaranteed to get one of them back from lookup.
\begin{align*}
  \text{parse}(\text{Trait}(\etag{t}, rec), r) = \\
  \text{parseTy}(rec[\text{lookup}(\etag{t}, r)], r)
\end{align*}
The final case for parse simply matches the value in the response to an index
into the record, and interprets that record entry using parseTy.

We will not write out the full definition of parseTy, but note that it is fairly
automatic. If the type is another declaration, we can call back to parse, if it
is a literal we simply generate the appropriate string, and if it is a record,
we build the appropriate record and recursively build out its fields.

\subsection{Correctness}
Given a declaration $d$, we generate a configuration, cfg. Then,
\begin{flalign*}
  &\forall\ (r : \textsf{Resp}).\\
  &r\ \text{well-formed}_{\text{cfg}} \Rightarrow \text{parse}(d, r)\ \text{defined}
\end{flalign*}
Note that since parse is a dependent function, it will already return a value of
the correct type. Thus, we only need to care that it is defined. Proof of this
fact is clear by induction on $d$.

@@

\subsection{Configuration Tool}
When a user wants to configure an Opal application to interact with an NLU
library like Wit, they begin by using our configuration tool. The tool turns
code in the type DSL into typescript code and JSON configuration, and ensures
that the NLU engine responses will agree with the type definitions in the user's
application.

@@

\section{Conclusion} \label{conclusion}
% TODO: Write

>>>>

\newcommand{\ff}[1]{\textsf{#1}}

@@

Finally, writing specific entity declarations rather than just \ts{type}, makes
it clear that there is more going on than a direct translation into Typescript.
In fact, the DSL parser does lot of validation to guarantee that only reasonable
types can be written down. For example, a \ts{keywords} declaration is required
to be a union, and that union must contain only literals. Even the \ts{alias}
keyword has some validation built in. Some Typescript types would be problematic
if they were allowed---we enforce those conditions in the DSL.

@@

We elected to use a Typescript library called \emph{runtypes} to serialize the
AST directly into Typescript; this gives us a data structure to guide response
parsing, and also provides us a way to dynamically type-check the results. We
discuss the response parsing algorithm in more detail in section
\ref{implementation}.

@@

\section{Formalism} \label{formalism}
In order to reason about the correctness of our translations, we have formalized
the results from section \ref{description}. We begin by formalizing the type DSL
as shown in figure \ref{fig:grammar_a}.
\begin{figure}
  \centering
  \begin{subfigure}{1\linewidth}
    \begin{align*}
      \fcy{D} ::=&\ \text{FreeText}(\etag{t}) \\
      |&\ \text{Keywords}(\etag{t},\ [\ell]) \\
      |&\ \text{Trait}(\etag{t},\ \{s: \fcy{T}\}) \\
      \\
      \fcy{T} ::=&\ \fcy{D} \mid \ell \mid \{s: \fcy{T}\}
    \end{align*}
    $$ \etag{t} \in \ff{Tag} \qquad \ell, s \in \textbf{string} $$
    \caption{Type DSL}
    \label{fig:grammar_a}
  \end{subfigure}\vspace{1cm}
  \begin{subfigure}{1\linewidth}
    $$ \tau ::=\ \textbf{string} \mid \ell \mid \{s: \tau\} \mid \langle \ell
    \rangle $$
    \caption{Relevant Opal Types}
    \label{fig:grammar_b}
  \end{subfigure}
  \caption{Formal grammars.}
\end{figure}
There are similarities between these definitions and the Haskell ones in figure
\ref{fig:ast}, but these have some slightly different goals. First, we remove
aliases and require that all type definitions are written ``in-line''. Next,
rather than split the definition by allowing tags within $\fcy{T}$, we also
require that entitiy definitions are inlined. This means that the data a user
expects from the NLU backend can be represented by a single tree of type
$\fcy{D}$. Importantly, none of these simplifications are fundamental---this
formulation can represent the same types as the actual system. This approach
simply gives up some usability to simplify reasoning.

@@

As mentioned above, user's view of their system can be formalized as a top-level
declaration generated by $\fcy{D}$. Since this declaration encompases both type
and configuration information, we would like a way to compute just the
\emph{type} of data that the user expects from the NLU engine. We define the
following mutually recursive functions, which traverse a declaration and
construct a single type in $\tau$.
\begin{align*}
  \ff{interp}\ &:\ \fcy{D} \to \tau \\
  \ff{interpTy}\ &:\ \fcy{T} \to \tau
\end{align*}

@@

\subsection{Configuration Generation}
The first step that our system takes is configuration. In practice, this is a
two-part process: the user needs to extract the appropriate Opal code as well as
the configuration for the NLU backend. The first step is covered by our
definition of \ff{interp}. Since we are taking $\tau$ as an analog for relevant
Opal types, the type generation process is exactly the execution of \ff{interp}
on the top-level entity declaration.

As for the second step, we would like the NLU backend to be set up such that its
responses are \emph{well-formed} with respect to the configuration. Intuitively,
any time we look up some entity tag in the response, we would like to get data
that is compatible with that entity's declaration. We define a type,
\textsf{Resp} of responses, and require that a \ff{lookup} function exists, with
the following type:
$$ \ff{lookup}\ :\ \ff{Tag} \to \ff{Resp} \rightharpoonup \textbf{string} $$
Note that the function is partial; there is no guarantee that the response will
contain the desired data. In the sequel, we will often assume that \ff{lookup}
is defined on tags we care about, but that assumption will be stated explicitly
rather than assumed by default.

The \ff{lookup} function always returns a string, but the string might mean any
number of things. The data might be a particular literal, in the case of a
keywords or trait entity, or an entire phrase in the case of a free-text entity.
We define configurations as functions that map a subset of entity tags to the
type of data expected for those entities.
$$ \ff{Cfg} \equiv \ff{Tag} \rightharpoonup \tau $$
Now, we call a response well-formed with respect to $\ff{cfg} : \ff{Cfg}$ if and
only if it returns appropriate data on \ff{lookup}. More formally, for some $r :
\ff{Resp}$,
\begin{align*}
  &r\ \text{well-formed}_{\text{cfg}} \iff \\
  &\forall (\etag{t} : \textsf{Tag}).\ \ff{cfg}(\etag{t})\ \text{defined} \Rightarrow \ff{lookup}(\etag{t}, r) : \ff{cfg}(\etag{t})
\end{align*}

@@

Our goal will be to write a parsing function which takes a declaration, $d$, and
a response, and produces a value of type \ff{interp}($d$). We will again write
two mutually recursive functions, \ff{parse} and \ff{parseTy}, with types:
\begin{align*}
\ff{parse}\ &:\ \Pi_{(d : \fcy{D})}. \textsf{Resp} \rightharpoonup \ff{interp}(d) \\
\ff{parseTy}\ &:\ \Pi_{(t : \fcy{T})}. \textsf{Resp} \rightharpoonup \ff{interp}(t)
\end{align*}
Since we need the type of the result to depend on the declaration, these are
dependent functions. The functions are also partial because lookups might fail.

We will define \ff{parse} by case analysis on the declaration.
\begin{align*}
  \ff{parse}(\text{FreeText}(\etag{t}), r) = \ff{lookup}(\etag{t}, r)
\end{align*}
This case is fairly straightforward. Since the interpretation of a free-text
entity is just a {\bf string}, we can simply look up the value of the
appropriate entity from the response.
\begin{align*}
  \ff{parse}(\text{Keywords}(\etag{t}, [\ell_1, \dots, \ell_n]), r) =\\
  \ff{lookup}(\etag{t}, r)
\end{align*}
Perhaps surprisingly, we can use the exact same definition in the Keywords case
that we did in the FreeText case. We are required to return a value of type
$\langle {\ell_1 \mid \dots \mid \ell_n} \rangle$; as long as $r$ is
well-formed, it will have been configured to have the same list of literals. We
are guaranteed to get one of them back from \ff{lookup}.
\begin{align*}
  \ff{parse}(\text{Trait}(\etag{t}, rec), r) = \\
  \ff{parseTy}(rec[\ff{lookup}(\etag{t}, r)], r)
\end{align*}
The final case for \ff{parse} simply matches the value in the response to an
index into the record, and interprets that record entry using \ff{parseTy}.

We will not write out the full definition of \ff{parseTy}, but note that it is
fairly automatic. If the type is another declaration, we can call back to
\ff{parse}, if it is a literal we simply generate the appropriate string, and if
it is a record, we build the appropriate record and recursively build out its
fields.

\subsection{Correctness}
Given a declaration $d$, we generate a configuration, \ff{cfg}. Then,
\begin{flalign*}
  &\forall\ (r : \textsf{Resp}).\\
  &r\ \text{well-formed}_{\text{cfg}} \Rightarrow \text{parse}(d, r)\ \text{defined}
\end{flalign*}
Note that since \ff{parse} is a dependent function, it will already return a
value of the correct type. Thus, we only need to care that it is defined. Proof
of this fact is clear by induction on $d$.

@@

\subsection{Configuration Tool}
When a user wants to configure an Opal application to interact with an NLU
library like Wit, they begin by using our configuration tool. The tool turns
code in the type DSL into Typescript code and JSON configuration, and ensures
that the NLU engine responses will agree with the type definitions in the user's
application.

@@

\section{Conclusion} \label{conclusion}
The Opal langauge hopes to be the ideal language for building modern
applications that leverage machine learning techniques. As NLU becomes more and
more popular as a user interface, it makes sense that Opal would provide a clean
and safe interface for natural langauge interaction.
